---
title: "Topic 1 - First steps"
author: "by Vianney Denis"
output: html_document
---

\INSTALL

### Download and install R
Go to [link](http://cran.r-project.org/bin/windows/base/) for windows system or [link](https://cran.r-project.org/) for mac. 
Run the installer, and follow the instructions, untick *'Save version number in registry'* 

\UPDATE

### Check for update

```{r,  eval=FALSE}
if(!require(installr)) {
  install.packages("installr"); require(installr)} 
updateR()
```

This will start the updating process of your R installation.  It will check for newer versions, and if one is available, will guide you through the decisions you need to make 

<span style="color: red"> It will tell you if your R version is out-of-date </span>

\PACKAGE

### Install and use a package

+ Install the package </span> <span style="color: blue">```abc```</span>

```{r,  eval=FALSE}
install.packages("abc")
```

<span style="color: red"> It will tell you if your package is updated </span>

To do only once, unless you remove or change of version of R 

+ Use the package <span style="color: blue">```abc```</span>

```{r,  eval=FALSE}
library("abc")
```

To do everytime: load your package 

> *<span style="color: green">**R.practice 1**: Install and load the package </span>* <span style="color: blue">```vegan```</span>

```{r,  eval=FALSE}
install.packages('vegan')
library('vegan')
```

\HELP

### Calling for help

```{r,  eval=FALSE}
?median
# or
help(median)
```

Help on the function <span style="color: blue">```median```</span>

```{r,  eval=FALSE}
?median
```

It will give you details on how to use the function <span style="color: blue">```median```</span>

```{r,  eval=FALSE}
??median
```
It will give you all functions with median in their description 
\WD

Use of working directory

```{r,  eval=FALSE}
getwd()
#Get Working Directory
```

Get the location of your current working directory

```{r,  eval=FALSE}
setwd()
#Set Working Directory
```

Set up a new working directory

\QUITR

### Quit R

```{r,  eval=FALSE}
q()
```
It will leave your R session, do not save your workspace 

\OBJECTS

### List objects

```{r,  eval=FALSE}
ls()
# list
```

It will list all objects in memory

```{r,  eval=FALSE}
rm(list=ls())
# remove list
```

It will remove all objects in memory

**Tip**: Use CTRL+L to clean your screen

\RCALCULATOR

### R is a calculator

```{r,  eval=FALSE}
3+2 # addition
3-2 # substraction
3*2 # multiplication
3/2 # division
3^3 # power
log(2) # logarithm
exp(2) # exponential
(5 + 3) / 4 # define priority using () or {} 
pi*4 # common function
```

R is case sensitive <span style="color: blue">```pi```</span> will work, whereas <span style="color: blue">```Pi```</span> will not </span> 

\EDITOR

### Text editor, Rite
Install and load package <span style="color: blue">```rite```</span> 

```{r,  eval=FALSE}
install.packages('rite')
library('rite')
rite()
riteout() # CTRL+Enter to send your lines
```

This is simple text editor

\READINGDATA

### Reading data
Using <span style="color: blue">```riteout```</span>

```{r,  eval=FALSE}
riteout() # CTRL+Enter to send your lines
setwd ('D:/.../Class 1')# set your working directory
read.table ("taiwan_coral.txt",header = TRUE,sep="\t",dec=".")# opening a simple datasheet
```

<span style="color: #e1d9d7"> You can also target directly your file without setting up your working directory </span>

```{r,  eval=FALSE}
read.table ("D:/.../Class 1/taiwan.txt",header = TRUE,sep="\t", dec=".")# long version
read.table("taiwan.txt", TRUE, "\t",".")# you can also shorten some part once you get use to it, but be careful when using this
```

<span style="color: #e1d9d7"> Now you can store it in R using the operator</span> <span style="color: #a2c6e5">```<-```</span><span style="color: #e1d9d7">or</span> <span style="color: #a2c6e5">```=```</span>

```{r,  eval=FALSE}
taiwan <- read.table ("taiwan.txt", 	header = TRUE, sep="\t", dec=".")
```
<span style="color: #a2c6e5">```read.delim```</span>
<span style="color: #e1d9d7">or</span>
<span style="color: #a2c6e5">```read.csv```</span> 
<span style="color: #e1d9d7"> and many others are alternative to</span>
<span style="color: #a2c6e5">```read.table```</span>  

<span style="color: #e1d9d7">Now, you can call your data by calling your object</span>
<span style="color: #a2c6e5">```taiwan```</span> 

```{r,  eval=FALSE}
taiwan
```
<span style="color: #e1d9d7">CONGRATS, this is your first script in R</span>


####__*<span style="color: green">Exercice 1b:</span>*__ <span style="color: green">using the script editor of R Studio ("File>New File>R Script""):</span>

#### *<span style="color: green">- change your working directory to the one you created for this class</span>*

#### *<span style="color: green">- import the file <span style="color: #a2c6e5">```rairuoho.txt```</span> (see below) in an object called rairuoho </span>*

*<span style="color: green">extra: try importing the url of the file using:<span style="color:#a2c6e5">```read.table()```</span>*

#### *<span style="color: green">- using the help facilities, find out, how you could read just the first lines, if needed (using the <span style="color: #a2c6e5">```read.table()```</span> function) </span>*

#### *<span style="color: green">- <span style="color: #a2c6e5">```sum()```</span>for this table, values from the 2nd column <span style="color: #a2c6e5">```rairuoho```</span><span style="color: #a2c6e5">```[```</span> no. line<span style="color: #a2c6e5">```,```</span> no. column<span style="color: #a2c6e5">```]```</span> </span>*


####<span style="color: green"> **Knit** your script ("File>Knit Document") in <span style="color: #a2c6e5">`html`</span> format and upload it below</span>
