---
title: "Fun with R:  Ecological data analysis in R"
author: "Vianney Denis"
date: "2020/09/15"
output:  
  "html_document":
   toc: true
   toc_depth: 3
   toc_float: true
   number_sections: false
   code_folding: show
    # ioslides_presentation 
---

# **Topic 1 - First steps** {.tabset}

## Introduction

### Syllabus

**Instructor:** Vianney DENIS [vianney.denis@gmail.com] - Functional Reef Ecology Laboratory, R406, 4F, Institute of Oceanography, National Taiwan University, Taiwan

**Students:** 12 students (pending, maybe less after today's class)

No.|Name | Major
---|-----|------
1|沈敬家|生科系(NTNU) 
2|Ya-Fang CHENG|海洋所博士班_一年級 
3|Ling-Lan HSU|海洋所博士班_二年級 
4|Yen-Ting CHEN|海洋所生漁組碩士班_一年級 
5|Tsai-Hsuan HSU|海洋所生漁組碩士班_一年級 
6|Yu-Wen CHEN|海洋所生漁組碩士班_一年級 
7|Che-Yueh LIN|海洋所生漁組碩士班_一年級
8|Kong-Mei SHUET|海洋所生漁組碩士班_一年級 
9|Wu-Chiao LIAO|海洋所生漁組碩士班_二年級 
10|Mariana REYES|氣候永續學程碩士班_二年級 
11|Chi-Le KO|生態演化所碩士班_一年級 
12|San-Pei LEE|生態演化所碩士班_二年級  


**Aim of the Course:**The aim of this course is to get familiar with the programming language R while avoiding technical jargon. We will explore together the advantages of using R language for formatting reports,  preparing presentations, exploring and analyzing ecoloical, data creating and sharing project. We will start from scratch, i.e. downloading R and installing it on your computer. Along this class, you will be introduced to  RStudio GUI, Rmarkdown, and GitHub repositories that complete harmoniously the use of R. 

At the end of this class, student should: 

- not be afraid of coding anymore and see the advantages in doing it;
- be self-learner, able to explore and solve problem;
- analyze and valorize scientifically a dataset;
- know several statistical tools for the analysis of ecological data in particular;
- help and recommend on the use of R.

**Class Philosophy:** I am not a statistician and I am not a programmer, so I will speak with simple words. My aim is to teach you bases in R language for you to be explore its power by yourself. The aim is not to know everything, but to be able to learn anything.

**Evaluation & Final Assignment:** Each week students prepare an example of the application of the things learned in previous class . One student randomly chosen will introduce its example at the beginning the class (using Rmarkdown). 

For the final assignment, group (2-3 students) will pick up a research topic along the semester, and will prepare a report emphasis methodology and results (scientific forrmat). Introduction and discussion can be presented as bullet points. Report should be sent as a markdown file before the last class of the semester. Do not wait the last minute!

---

**Disclaimer:** Format of course largely inspired by documentation from Fukami T. & Coyle J. (Standford University). I am currently reshaping all documents to have more interaction with students during this course. 

### Tentative Schedule 

Schedule is highly susceptible to changes as I am updating content every week.

| Week | Day | Date     | Tentative Topic | Content|
|------|-----|----------|-----------------|--------|
| 1    | Tue | ---------| Introduction: First step | Introduction R, R Studio, Markdown, GitHub|
| 2 & 3 | Tue | ---------| Data manipulation, basic graphics and statistical functions | Manipulate dataset, basic operation, make and edit a plot|
| 4 & 5 | Tue | ---------| Data exploration | Summarize dataset, preliminary analyses |
| 6 | Tue | ---------| Ecological resemblance | Association coefficient, similarity matrix |
| 7 | Tue | ---------| Numerical classification | Cluster analysis |
| 8 & 9 | Tue | ---------| Unconstrained ordination | PCA, PCoA, nMDS |
| 10 & 11 | Tue | ---------| Constrained ordination | Redundancy and cannonical analysis |
| 12 & 13 | Tue | ---------| Generalized Linear Model | Linear regression, and GLMs |
| 14 | Tue | ---------| From traits to functions | --- |

### References & Readings

Borcard, D., Gillet, F. and Legendre, P. (2018) Numerical Ecology with R. Springer. DOI: [10.1007/978-3-319-71404-2](https://www.springer.com/gp/book/9783319714035)


Paradis, E. (2005) R for beginner. Available among many others documentation at https://cran.r-project.org/other-docs.html (page frozen and no longer actively maintained) 

Ramette, A. (2007) Multivariate analyses in microbial ecology. FEMS Microbiology Ecology, 62: 142-160. DOI: [10.1111/j.1574-6941.2007.00375.x](https://onlinelibrary.wiley.com/doi/full/10.1111/j.1574-6941.2007.00375.x)

Xie, Y., Allaire, J.J.,Grolemund, G. (2020) R Markdown: The Definitive Guide. Chapman & Hall/CRC. The online version of this book is free to read [here](https://bookdown.org/yihui/rmarkdown/). 

Zuur, A. F., Ieno E. N., Smith, G. M. (2007) Analysing Ecological Data. Springer.  DOI: [10.1007/978-0-387-45972-1](https://www.springer.com/gp/book/9780387459677)

Zuur, A. F. (2010) A protocol for data exploration to avoid common statistical problems. Methods in Ecology and Evolution, 1: 3-14. DOI: [10.1111/j.2041-210X.2009.00001.x](https://besjournals.onlinelibrary.wiley.com/doi/10.1111/j.2041-210X.2009.00001.x)

## R software

### What's R?

R is a popular free and open program for statistical computing (https://www.r-project.org/about.html). R is a widely-used program for statistical computing. R can be run from the command line or from a graphical user interface (GUI) that is automatically installed when you install R (if you open the R program installed on your computer, this is what will open up). 

R consists of a core (minimal installation + add on packages (from the Comprehensive R Archive Network, Bioconductor, or others). The core's capabilities is expanded by using the add-on packages.

### Installation

#### Windows

If you don't have R installed on your computer:

- Download R from the [CRAN website](https://cran.r-project.org/bin/windows/base/release.htm).
- Run the `.exe` file that was just downloaded.
- Once it’s installed, open R to make sure it works and you don’t get any error messages.

If you already have R already install:

To check the version of R you are using, start R: the version of R you are running is the first thing that appears on the terminal. Alternatively, Type `sessionInfo()` in the R console, which will  display which version of R you are running. Go on the [CRAN website](https://cran.r-project.org/) and check whether a more recent version is available. If so, please download and install it. You can check [here](https://cran.r-project.org/bin/windows/base/rw-FAQ.html#How-do-I-UNinstall-R_003f) for more information on how to remove old versions from your system if you wish to do so.

#### macOS

If you don't have R installed on your computer:

- Download R from the [CRAN website](https://cran.r-project.org/bin/macosx/).
- Select the `.pkg` file for the latest R version.
- Double click on the downloaded file to install R
- It is also a good idea to install [XQuartz] (https://www.xquartz.org/) (needed by some packages)
- Once it’s installed, open R to make sure it works and you don’t get any error messages.

If you already have R already install:

To check the version of R you are using, start R: the version of R you are running is the first thing that appears on the terminal. Alternatively, Type `sessionInfo()` in the R console, which will  display which version of R you are running. Go on the [CRAN website](https://cran.r-project.org/bin/macosx/) and check whether a more recent version is available. If so, please download and install it. 


#### Linux

- Follow the instructions for your distribution from [CRAN](https://cloud.r-project.org/bin/linux/), they provide information to get the most recent version of R for common distributions. 

- Once it’s installed, open R to make sure it works and you don’t get any error messages.


### Update

If you already have R on your computer, an alternative is to just run (<span style="color: red">it will tell you if your R version is out-of-date </span>): 

```{r,  eval=FALSE, tidy=TRUE}
if(!require(installr)) {
  install.packages("installr"); require(installr)} 
updateR()
```

This will start the updating process of your R installation.  It will check for newer versions, and if one is available, will guide you through the decisions you need to make. 

You may have to choose a CRAN mirror if it is your first download. This is just a copy of the CRAN archives. Choose one in Taiwan, it is usually a bit faster to download and install. 

### Package

To expand the abilities of R, you can first install a package with its name. The package `ABC` provides tools for Approximate Bayesian Computation (ABC) 

- Install the package `abc`

```{r,  eval=FALSE}
install.packages("abc")
```

```{r,  eval=FALSE}
# all information after `#` is ignored, this is VERY important and use to comment your code.
```

```{r,  eval=FALSE}
install.packages("abc") # I am installing the package 'abc'
```

- Use the package `abc`

```{r,  eval=FALSE}
library("abc")
```

To do every time you wanna use the package: load your package 


> *<span style="color: green">**RP1**: Install and load the package </span>* `vegan`

```{r class.source = "fold-hide",  eval=FALSE}
install.packages('vegan')
library('vegan')
```

If you don't know the name of the package: 

- use task views available at http://cran.r-project.org/web/views that group packages to subject areas such environmetrics, multivariate, etc. 

- many online discussion, forum, and docs. Among many others sources, you will often be directed to discusion in [stackoverflow](https://stackoverflow.com/)

### Help

You can acces the html help via the Help menu: *Help/Search help*. Alternatively, a faster solution is to call for help directly in the console:

- a quick summary on how to use the function `median`

```{r class.source = "fold-show",  eval=FALSE}
?median
```

- a detailed summary on the use of the function `median```</span>

```{r class.source = "fold-show",  eval=FALSE}
?median
# or
help(median)
```

- it will give you all functions with median in their description 

```{r,  eval=FALSE}
??median
```

### Working directory

It is a good practice to create a working directory (folder) where you will store all your file. In general, each project will have its working directory including data set, script, plot, etc

Use of working directory. Retrieve it using:

```{r,  eval=FALSE}
getwd() #get the location of your working directory
```

Set it up using:

```{r,  eval=FALSE}
setwd()
#get the location of your working directory
```

### Quit R

**File -> Exit** or use the console:

```{r,  eval=FALSE}
q()
```

Saving the workspace image will allow you to save all objects created in the R memory into your working directory (.Rdata). It also saves all commands given during the session (.Rhistory).

Quit without saving for now.

### List

`ls()` will list all the objects in the memory.

```{r,  eval=FALSE}
a<-'corals' # create an object 'a' containing 'corals'
b<-'are' # create an object 'b' containing 'are'
c<-'cool'# create an object 'c' containing 'cool'
ls # list objects 'a', 'b', 'c'
```

You can clean the objects present in the memory using:

```{r,  eval=FALSE}
rm(list=ls()) # rm: remove ls: list of all objects present in memory
```

**Tip**: Use CTRL+L to clean your screen when you want to make it clear. It does not remove the objects in memory.

### Calculator

R is a calculator:

```{r,  eval=FALSE}
3+2 # addition
3-2 # substraction
3*2 # multiplication
3/2 # division
3^3 # power
log(2) # logarithm
exp(2) # exponential
(5 + 3) / 4 # define priority using () or {} 
pi*4 # common function
```

R is case sensitive <span style="color: blue">```pi```</span> will work, whereas <span style="color: blue">```Pi```</span> will not </span> 

### Reading

To be able to work with a dataset (and make calculation based on it), it has to be import as an object in the memory of R. Many formats can be used (.xls, .spss, SAS, etc.). To start with, we will use and .xlsx file, but quickly we will turn toward less heavy format. Note the distinction between **reading** a file and **importing** a file.

My file `reef_fish` is an .xlsx file located in the file 'Data' of my working directory. 

```{r,  eval=FALSE}
library(readxl) # I am loading a package (library) allowing me to read .xls file
read_excel('Data/reef_fish.xlsx') # I am reading my file
fish<-read_excel('Data/reef_fish.xlsx') # I am importing my file in an object called 'fish'
```

More commonly, we will use .txt file (lighter) and the function `read.table'. But often you will need to think at several 'arguments':

- What is the name of the data file (with the extension)?
- Does the file have column titles?
- What is the delimiter between the columns (tab, space, comma or semicolon)?
- What is the character used for decimal
points (dot, comma)?

```{r,  eval=FALSE}
fish<-read.table('Data/reef_fish.txt', header=T, sep='\t', dec='.') # reading txt file
```

Check `?read.table` for all arguments you can specify within this function

Instead or working in  a working directory, you can also target directly your file (check your file properties) or  an url.

```{r,  eval=FALSE,  tidy=TRUE}
fish<-read.table ("D:/.../Topic 1/Data/reef_fish.txt",header = TRUE,sep="\t", dec=".")# long version
fish<-read.table("taiwan.txt", TRUE, "\t",".")# you can also shorten some part once you get use to it, but be careful when using this
```

You can also use `file.choose()' to locate dircetly the file or or computer or alternatively import data from the R menu.

```{r,  eval=FALSE,  tidy=TRUE}
fish<-read.table(file.choose (), header = TRUE,sep="\t", dec=".")
```

**Note_1**: in all cases creation of a file with the same name 'fish' will replace the one previously created.

**Note_2**: two operators `<-` and `=` can be used to import data. While longer, I favor `<-` in the creation of object to avoid confusion with mathematical operator. 

### Editor

You can feel it will be interesting to have an editor for our code. It will help at saving and running quickly our code in R. Many alternative: open a .txt file, open an new script in R, use a text editor. The latest one is the most commone altrenative.


## RStudio

### What's RStudio?

RStudio is a graphical integrated development environment (IDE) that makes using R much easier and more interactive. It is open source (i.e. free)
and available at http://www.rstudio.com/

### Installation

If you don't have RStudio installed on your computer:

- Go to the [RStudio download page](https://rstudio.com/products/rstudio/download/#download).
- Under Installers select RStudio x.yy.zzz - Windows XP/Vista/7/8 (where x, y, and z represent version numbers)
- Double click the file to install it
- Once it’s installed, open RStudio to make sure it works and you don’t get any error messages

If you already have R already install:

- Open RStudio, and click on “Help” > “Check for updates”. If a new version is available, quit RStudio, and download the latest version for RStudio

### Environment

RStudio combines (File>New File>R Script):

- A script editor window where you can edit and send you commands to the console

- The **Console** is where you can type commands and see the output

- The **Workspace** shows all the active objects. The history tab shows a
list of commands used so far.

- The **Files** tab shows all the files and folder in your default workspace. The **Plots** tab will show all your graphs. The **Packages** tab will list a series of packages or add-ons needed to run certain processes.
For additional info see the **Help** tab.

Among the many advantages of RStudio, you will appreciate the user-adjustable interface (see Options>Appearance>Layout), the auto-synthax, the suggestion `function {package}` together a brief description of its use.


Using our `fish` dataset, write a code to import data as an object `fish` then copy the following 'plot' command in in your script. Run the script, and observe the result.

```{r,  eval=F}
fish<-read.table('Data/reef_fish.txt', header=T, sep='\t', dec='.')
barplot(fish$richness, main="Top 10 reef fish Richness (Allen, 2000)", horiz=TRUE, names.arg=fish$country, cex.names=0.5, las=1)
```

**Note**: you can easily export the resulting plot as and image, a pdf, or to copy it to your clipboard.  

## Documentation

Our course is a set of R markdown files that have been rendered to html files and are hosted on a repository https://github.com/vianneydenis/attentive-suricate.git. 

Each Topic of the class has a corresponding R markdown file named Topic_X_topic.Rmd, where X is the topic number. Today for example the file Topic_1_Intro.Rmd allows you to play and review the class today. You can render today .Rmd by 'kniting' it in RStudio. Options allow you to 'Preview in Window' or to 'Preview in viewer Panel'

This repository working directory available on GitHub is the working directory of the attentive-suricate.Rproj file and it contains all of the .Rmd, .html and data files used to review this course. While you can can individually download and render the .Rmd files. It is better to 'fork' this repository to your GitHub account, then clone a copy of this forked repository  (from your own Github account to your computer). This only needs to be done once. 

### GitHub

Create an account on [GitHub](https://github.com/). You can **create your first repository** to later connect it into  RStudio.

### Git

Install Git on your computer

#### Windows

1. Download the git installer for Windows [here](https://git-scm.com/). This will install bash, which is a command-line shell program for interacting with files and programs on your computer.

2. Run the installer that you just downloaded. Click “Next” as needed and when you get to the following options accept default.

3. That’s it! You should now have three new programs installed in a folder called Git under your Program Files.

#### macOS

Follow the instructions below or view this [video tutorial](https://www.youtube.com/watch?v=9LQhwETCdwY&ab_channel=SarahStevens).

1. Go to this [list](https://sourceforge.net/projects/git-osx-installer/files/) and download an installer for Git.
   + If you are using OS X 10.9 and higher, choose the most recent version for “mavericks”.
   + If you are using an earlier version of OS X, choose the most recent version for “snow leopard”.

2. Run the installer you downloaded.

3. After installing Git, there will not be anything in your /Applications folder, because Git is a command line program.


**Note**: if you have OSX El Capitan you may experience some errors when running Git if you don’t have an updated XCODE.[Here](https://stackoverflow.com/questions/32893412/command-line-tools-not-working-os-x-el-capitan-sierra-high-sierra-mojave) is a description of the issue on StackOverflow and the solution is to open a Terminal window and type `xcode-select--install`.

#### Linux 

You may already have Git installed. If not, you can try to install it via your dsitribution package manager. For Debian/Ubuntu run 'sudo apt-get install git' and for Fedora run 'sudo yum install git'.


### RStudio

File > New Project > Version Control 'Checkout a project from a version control repository'. Enter URL of your repository, (its name), and where you want to copy it on your computer. Once you validate:

+ you will create a .RProject. Any change in your repository will appear in 'Git'
+ changes needs to be 'commit' (prepared) and 'push' (uploaded) to GitHub in order to be synchronized
+ first 'commit' will fail and ask you to set up Git with email and username. Open Git CMD and process as indicated
+ first 'push' will be pending your GitHub login and password
+ any changes in your GitHub repo must be 'pull' prior working on your project in RStudio

### Rmarkdown

R markdown is text-based formatting that allows you to embed code and explanatory text in the same document.R markdown documents consist of a header, code chunks and text. R markdown files (.Rmd) can be rendered to other document formats (e.g. html, pdf, docx) to generate reports or web applications. For example, the present document was generated as an html from a Rmd file. Formatting analyses in R markdown keeps the entire research process in one document that can easily regenerate reports and output if the data are changed.

Visit R markdown [webpage](https://rmarkdown.rstudio.com/) for an overview of its capability. [Download](https://github.com/rstudio/cheatsheets/raw/master/rmarkdown-2.0.pdf) cheat sheet  to get familiar with it. I also recommend you R markdown bible: Xie et al. (2020), available [here](https://bookdown.org/yihui/rmarkdown/) to become an expert with markdown. 


As for now, just try in RStudio: 

*File > New File > R markdown* in order to create a simple .html document. Try to modify and play with it with the information available in the cheat sheets you downloaded.  

> Attribution
>
> The content of this webpage was derived from Fukami T. & Coyle J. (Standford University) which themselves derived from Data Carpentry and Software Carpentry materials (Copyright (c) Data Carpentry) under CC-3.0 Attribution License (see http://creativecommons.org/licenses/by/3.0/).

## Assignements

You have two main missions along the semester:

- Final project: research projects (2-3 students) must be ready  **one** week before the end of the semester. A final report (.Rmd) will be send to the instructor 1 week before the final, and students will present their research on our last week together. The report will follow the format of a scientific publication but the style of the introduction and discussion can used bullet point to emphasize the methodology and the results. Topic is free, but a small proposal (1-2 pages, .Rmd) must be submitted before mid-term. Mid-term week will be dedicated to a discussion of the **research project**. Topics are free (not necessarily restricted to Ecology), and you are welcome to develop your own experiment but you can also use online published datasets which you will reinterpret. 



***
**Examples of topics for the final assignment:**



- Diversity and distribution of birds on NTU campus

*This project aims at documenting the diversity of birds on NTU campus and at examining their habitat preferences. Overall, this study proposes to appreciate the role of the campus for the avian fauna in a megacity such Taipei*

- YouBike 2.0 users: gender, age, occupation. A demographic studies.

*This project aims at uncovering who are the users of YouBike 2.0 and if differences exist among genre, age and occupation. This study will contribute at a greater understanding on people circulate on the campus*


- Diel variation in temperature, pH, and oxygen in water from the Drunken Moon Lake: a correlative approach

*This project will aim at uncovering diel variations in several key physiological and biological parameters of the Drunken Moon Lake. This project will aim at a better characterization of the environmental in this artificial lake*

Etc.


- After rendering course file in RStudio, you can review it. Each  week, you will prepare a short example based on what you learned in the previous class - or how to extended it. At the beginning of each class, a presenter will be randomly selected to illustrate what he prepared (5-10 minutes). The algorithm below will decide who will present. You may present several time in the semester (lucky) or none (unlucky), so always be ready. You will prepare your summary in .Rmd format

```{r,  eval=F}
library(dplyr)
list_student<-c('沈敬家','Ya-Fang CHENG','Ling-Lan HSU','Yen-Ting CHEN','Tsai-Hsuan HSU','Yu-Wen CHEN','Che-Yueh LIN','Kong-Mei SHUET','Wu-Chiao LIAO','Mariana REYES','Chi-Le KO','San-Pei LEE')
sample(list_student, 1, replace = TRUE)
```

Next week: install and load a package, set up your working directory, load a dataset and look for some help on the application of a given function


