---
title: "Fun with R:  Ecological data analysis in R"
author: "Vianney Denis"
date: "2020/10/20"
output:  
  "html_document":
   theme: united
   highlight: tango
   toc: true
   toc_depth: 4
   toc_float: true
   number_sections: false
   code_folding: show
    # ioslides_presentation 
---


# **Topic 4 - Make a map**

Current solutions for creating maps usually involves GIS software, such as ArcGIS, QGIS, eSpatial, etc., which allow to visually prepare a map, in the same approach as one would prepare a poster or a document layout. On the other hand, R has also advanced spatial capabilities and can be used to draw maps. As we mentioned earlier, you can customize easily a map as you wish.  

## `maptools` and `rgdal` packages

### Simple map 

Getting a simple map is pretty straightforward  in R using the package`maptools`

```{r,  eval=T, message=F}
library('maptools')
data(wrld_simpl)
plot(wrld_simpl)
```

Now we know the options to modify a plot, therefore you can customize this map with some elements we mentioned earlier:

```{r,  eval=T}
plot(wrld_simpl,xlim=c(100,130),ylim=c(-20,50),col='olivedrab3',bg='lightblue')
```

Try to zoom on Taiwan: you will quickly realize that we have a pretty bad resolution a finer scale, see `?wrld_simpl`  

### GPS: data tracking

If you like running, hiking, walking, or being lazy in a park - many devices now can export `gpx` format. You an read the file directly as lines (i.e, `tracks`), points (i.e, `track_points`), and a few other formats you can find in the help for `readOGR`. 

```{r,  eval=T,message=F, warning=F}
library(rgdal)
par(mfrow=c(1,2))
run1 <- readOGR(dsn="Data/run.gpx",layer="tracks")
plot(run1, main='Line') # my running activity
run2 <- readOGR(dsn="Data/run.gpx",layer="track_points")
plot(run2, main='Points')
dev.off()
```

The reverse can be accomplish with the function `writeOGR` to export  objects as SpatialPointsDataFrame, SpatialLinesDataFrame, or SpatialPolygonsDataFrame  (as defined in the `sp` package). This will write an ArcGIS compatible shapefile, but many different formats are available by specyfying the correct driver.Now that we know how to get a super basic map in R, let's look at how we can export and import data. 

```{r,  eval=T,message=F, warning=F}
writeOGR(wrld_simpl,dsn="Data", layer = "world_test", driver = "ESRI Shapefile", overwrite_layer = TRUE)
```

Now we could open `world_test.shp` in ArcGIS (or others), but we can also import shapefiles (.shp) back into R, let's use that same file

```{r,  eval=T,message=F, warning=F}
world_shp <- readOGR(dsn = "Data",layer = "world_test")
plot(world_shp)
```

## Spatial data types</span>

#### Vector based (points, Lines, polygones)

Creating spatial data from scratch in R seems a little bit confusing at first, but once you got the logic behind, it get easier:

- `SpatialPointsDataFrame` for plotting points

```{r,  eval=T}
plot(wrld_simpl,xlim=c(115,128) ,ylim=c(19.5,27.5),col='#D2B48C',bg='lightblue') # TW map
coords <- matrix(c(121.537290,120.265541, 25.021335, 22.626524),ncol=2) # NTU and SYS univ. 
coords <- coordinates(coords) # assign values as spatial coordinates
spoints <- SpatialPoints(coords) # create SpatialPoints
df <- data.frame(location=c("NTU","SYS")) # create a dataframe
spointsdf <- SpatialPointsDataFrame(spoints,df) # create a SpatialPointsDataFrame
plot(spointsdf,add=T,col=c('black','black'),pch=19,cex=2.2) # plot it on our map
text(121,24, 'TAIWAN', cex=1)
```

- `SpatialLinesDataFrame` for plotting lines: let's travel in Canada in the province of Saskatchewan 

```{r,  eval=T}
plot(wrld_simpl,xlim=c(-130,-60),ylim=c(45,80),col='#D2B48C',bg='lightblue')
coords <- matrix(c(-110,-102,-102,-110,-110,60,60,49,49,60),ncol=2)
l <- Line(coords)
ls <- Lines(list(l),ID="1")
sls <- SpatialLines(list(ls))
df <- data.frame(province="Saskatchewan")
sldf <- SpatialLinesDataFrame(sls,df)
plot(sldf,add=T,col='#3d2402', cex=2)
text(-114, 55, 'Saskatchewan', srt=90, cex=0.7)
text(-114, 63, 'CANADA', cex=1)
```

- `SpatialPolygonesDataFrame` for plotting polygones

```{r,  eval=T}
plot(wrld_simpl,xlim=c(-130,-60),ylim=c(45,80),col='#D2B48C',bg='lightblue')
coords <- matrix(c(-110,-102,-102,-110,-110,60,60,49,49,60),ncol=2)
p <- Polygon(coords)
ps <- Polygons(list(p),ID="1")
sps <- SpatialPolygons(list(ps))
df <- data.frame(province="Saskatchewan")
spdf <- SpatialPolygonsDataFrame(sps,df)
plot(spdf,add=T,col='#45220d') 
text(-114, 55, 'Saskatchewan', srt=90, cex=0.7)
text(-114, 63, 'CANADA', cex=1)
text(-103, 46, 'UNITED STATES', cex=1)
text(-40, 78, 'GREENLAND', cex=1)
text(-35, 55, 'Atlantic Ocean', cex=1, col='#071238')
```


## Map improvement: `raster`

Spatial data: download, unzip, and import spatial shape (don't forget to set up your working directory). Those spataial data are higher resolution spatial data at teh country level.

```{r,  eval=T}
download.file("https://www.dipintothereef.com/uploads/3/7/3/5/37359245/twn_adm.zip", destfile="Data/taiwan_shape.zip") # download file
unzip('taiwan_shape.zip',exdir="Data/taiwan_shape_unzip") # unzip it 
taiwan <- readOGR(dsn = "Data/taiwan_shape_unzip",layer = "TWN_adm0")
# can also sf::st_read in order to read data
```

Plot the map of the whole country:

```{r,  eval=T}
plot (taiwan) # simple naked map
plot (taiwan, axes=T, xlim=c(121,122), ylim=c(24,25.5), bg=colors()[431],col='grey') # add colors
```

Using the `raster`package, and the function getData`, you will download directly polygones (vectorized) shape from the GADM (Global Administrative Areas, or other sources). You will need country code: `TWN` for Taiwan, `JPN` for Japan, etc. The argument `level` indicates the level you are trying to get. For province level `level=1`, for the overall country level `level=1`

```{r,  eval=T}
library(raster)
TWN <- getData('GADM', country="TWN", level=1) # data Taiwan
JPN <- getData('GADM', country="JPN", level=1) # data Japan
plot(JPN,axes=T,bg=colors()[431],col='grey')
```

We can examined the object `TWN`

```{r,  eval=FALSE}
class(TWN) # SpatialPolygonsDataFrame
TWN$NAME_1 # gives the names of the provinces, incomplete in TW
```

And plot individual province as:

```{r,  eval=T}
plot(TWN,col="grey",xlim=c(119,122.5), ylim=c(21.5,25.5), bg=colors()[431], axes=T)
KAO <- TWN[TWN$NAME_1=="Kaohsiung",]
plot(KAO,col="gray 33",add=TRUE)
```

Random customization of our map (e.g. adding population)

```{r,  eval=T}
plot(TWN,col="grey",xlim=c(121,122), ylim=c(24,25.5), bg=colors()[431], axes=T)
coords <- matrix(cbind(lon=c(121.2,121.6,121.8),lat=c(25,25.1,24.5)),ncol=2)
coords <- coordinates(coords)
spoints <- SpatialPoints(coords)
df <- data.frame(location=c("City 1","City 2","City 3"),pop=c(138644,390095,34562))
spointsdf <- SpatialPointsDataFrame(spoints,df)
scalefactor <- sqrt(spointsdf$pop)/sqrt(max(spointsdf$pop))
plot(spointsdf,add=TRUE,col='black',pch=16,cex=scalefactor*5) 
# add location of NTU
points(121.537290,25.021335, type="p", pch=18, col=2, cex=2)
# add text
text(121.437290,24.921335,"NTU", col='red', font=2)
# add scale
maps::map.scale(x=121.8, y=24.15)
# add north arrow
GISTools::north.arrow(xb=122.25,yb=24.5, len=0.06, lab='N')
```


## Mapping with `ggplot2`and `sf`

As we mentioned earlier, the package `ggplot2` implements the grammar of graphics in R. While ggplot2 is becoming the standard for R graphs, it does not handle spatial data specifically. The current state-of-the-art of spatial objects in R relies on Spatial classes defined in the package `sp`, but the new package `sf` has recently implemented the “simple feature” standard, and is steadily taking over `sp`. Recently, the package `ggplot2` has allowed the use of simple features from the package `sf` as layers in a graph (since the version 3.0.0 of `ggplot2`). The combination of `ggplot2` and `sf` therefore enables to programmatically create maps, using the grammar of graphics, just as informative or visually appealing as traditional GIS software.

### Getting started: dataset


We will start by loading the basic packages necessary for all maps, i.e. `ggplot2` and `sf`. We also suggest to use the classic dark-on-light theme for `ggplot2` (`theme_bw`), which is appropriate for maps:

```{r,  eval=T}
library('ggplot2')
theme_set(theme_bw()) 
library('sf')
```

The package `rnaturalearth` provides a map of countries of the entire world. Use `ne_countries` to pull country data and choose the scale (`rnaturalearthhires` is necessary for `scale = "large"`). The function can return `sp` classes (default) or directly `sf` classes, as defined in the argument `returnclass`:

```{r,  eval=T}
library("rnaturalearth")
library("rnaturalearthdata")
world <- ne_countries(scale = "medium", returnclass = "sf")
class(world)
# wrld_simpl <- st_as_sf(wrld_simpl) for conversion from st to sf
```

### Data and basic plot (`ggplot` and `geom_sf`)

First, let us start with creating a base map of the world using `ggplot2`. This base map will then be extended with different map elements, as well as zoomed in to an area of interest. We can check that the world map was properly retrieved and converted into an `sf` object, and plot it with `ggplot2`

```{r,  eval=T}
ggplot(data = world) +
   geom_sf()
```

In your map of the world, the plot panel is expanded beyond the size of the earth (you can see that the graticule lines end before the edge of the plot panel), and hence no axis ticks are drawn. One way to solve the issue is to turn off the expansion.

```{r,  eval=T}
ggplot(data = world) +
   geom_sf() +
   coord_sf(expand = FALSE)
```


As before, the  layers are added one at a time in a `ggplot` call, so the order of each layer is very important. All data will have to be in an `sf` format to be used by `ggplot2`; data in other formats (e.g. classes from `sp`) will be manually converted to `sf` classes if necessary.

- Title, subtitle, and axis labels (`ggtitle`, `xlab`, `ylab`)

A title and a subtitle can be added to the map using the function `ggtitle`, passing any valid character string (e.g. with quotation marks) as arguments. Axis names are absent by default on a map, but can be changed to something more suitable (e.g. “Longitude” and “Latitude”), depending on the map:

```{r,  eval=T}
ggplot(data = world) +
    geom_sf() +
    coord_sf(expand = FALSE) +
    xlab("Longitude") + ylab("Latitude") +
    ggtitle("World map", subtitle = paste0("(", length(unique(world$name)), " countries)"))
```

- Map color (`geom_sf`)

In many ways, `sf` geometries are no different than regular geometries, and can be displayed with the same level of control on their attributes. Here is an example with the polygons of the countries filled with a green color (argument `fill`), using black for the outline of the countries (argument `color`):

```{r,  eval=T}
ggplot(data = world) + 
    geom_sf(color = "black", fill = "lightgreen") +
    coord_sf(expand = FALSE) 
```


The package `ggplot2` allows the use of more complex color schemes, such as a gradient on one variable of the data. Here is another example that shows the population of each country. In this example, we use the "viridis” colorblind-friendly palette for the color gradient (with `option = "magma"` for the magma variant), using the square root of the population (which is stored in the variable `POP_EST` of the `world` object):

```{r,  eval=T}
ggplot(data = world) +
    geom_sf(aes(fill = pop_est)) +
    coord_sf(expand = FALSE) +
    scale_fill_viridis_c(option = "plasma", trans = "sqrt")
```

- Projection and extend (`coord_sf`)

The function `coord_sf` allows to deal with the coordinate system, which includes both projection and extent of the map. By default, the map will use the coordinate system of the first layer that defines one (i.e. scanned in the order provided), or if none, fall back on WGS84 (latitude/longitude, the reference system used in GPS). Using the argument `crs`, it is possible to override this setting, and project on the fly to any projection. This can be achieved using any valid PROJ4 string (here, the European-centric ETRS89 Lambert Azimuthal Equal-Area projection):



```{r,  eval=T}
ggplot(data = world) +
    geom_sf() +
    scale_fill_viridis_c(option = "plasma", trans = "sqrt") +
    coord_sf(crs = "+proj=laea +lat_0=52 +lon_0=10 +x_0=4321000 +y_0=3210000 +ellps=GRS80 +units=m +no_defs ")
```

Spatial Reference System Identifier (SRID) or an European Petroleum Survey Group (EPSG) code are available for the projection of interest, they can be used directly instead of the full PROJ4 string. The two following calls are equivalent for the ETRS89 Lambert Azimuthal Equal-Area projection, which is EPSG code 3035:

```{r,  eval=F}
ggplot(data = world) +
    geom_sf() +
    coord_sf(crs = "+init=epsg:3035")

ggplot(data = world) +
    geom_sf() +
    coord_sf(crs = st_crs(3035))
```


The extent of the map can also be set in `coord_sf`, in practice allowing to “zoom” in the area of interest, provided by limits on the x-axis (`xlim`), and on the y-axis (`ylim`). Note that the limits are automatically expanded by a fraction to ensure that data and axes don’t overlap; it can also be turned off to exactly match the limits provided with `expand = FALSE`:

```{r,  eval=T}
ggplot(data = world) +
    geom_sf() +
    coord_sf(xlim = c(118, 128), ylim = c(17, 27), expand = FALSE)
```

- Scale bar and North arrow (package `ggspatial`)

Several packages are available to create a scale bar on a map (e.g. `prettymapr`, `vcd`, `ggsn`, or `legendMap`).Here the package `ggspatial` provides easy-to-use functions. `scale_bar` that allows to add simultaneously the north symbol and a scale bar into the `ggplot` map. Five arguments need to be set manually: `lon`, `lat`, `distance_lon`, `distance_lat`, and `distance_legend`. The location of the scale bar has to be specified in longitude/latitude in the `lon` and `lat` arguments. The shaded distance inside the scale bar is controlled by the `distance_lon` argument while its width is determined by `distance_lat`. Additionally, it is possible to change the font size for the legend of the scale bar (argument `legend_size`, which defaults to 3). The North arrow behind the “N” north symbol can also be adjusted for its length (`arrow_length`), its distance to the scale (`arrow_distance`), or the size the N north symbol itself (`arrow_north_size`, which defaults to 6). Note that all distances (`distance_lon`, `distance_lat`, `distance_legend`, `arrow_length`, `arrow_distance`) are set to `"km"` by default in `distance_unit`; they can also be set to nautical miles with `“nm”`, or miles with `“mi”`.

```{r,  eval=T}
library(ggspatial)
ggplot(data = world) +
    geom_sf() +
    annotation_scale(location = "br", width_hint = 0.4) +
    annotation_north_arrow(location = "br", which_north = "true", 
        pad_x = unit(0.5, "cm"), pad_y = unit(1, "cm"),
        style = north_arrow_fancy_orienteering) +
    coord_sf(xlim = c(118, 128), ylim = c(17, 27), expand = FALSE)
```

If you plot a larger area, you may get a warning on the inaccuracy of the scale bar.

- Country names and other names (geom_text and annotate)
The `world` data set already contains country names and the coordinates of the centroid of each country (among more information). We can use this information to plot country names, using `world` as a regular `data.frame` in `ggplot2`. The function `geom_text` can be used to add a layer of text to a map using geographic coordinates. The function requires the data needed to enter the country names, which is the same data as the world map. Again, we have a very flexible control to adjust the text at will on many aspects:
   + The size (argument `size`);
   + The alignment, which is centered by default on the coordinates provided. The text can be adjusted horizontally or vertically using the arguments `hjust` and `vjust`, which can either be a number between 0 (right/bottom) and 1 (top/left) or a character (“left”, “middle”, “right”, “bottom”, “center”, “top”). The text can also be offset horizontally or vertically with the argument `nudge_x` and `nudge_y`;
   + The font of the text, for instance its color (argument `color`) or the type of font (`fontface`);
   + The overlap of labels, using the argument `check_overlap`, which removes overlapping text. Alternatively, when there is a lot of overlapping labels, the package `ggrepel` provides a `geom_text_repel` function that moves label around so that they do not overlap.
   + For the text labels, we are defining the centroid of the countries with `st_centroid`, from the package `sf`. Then we combined the coordinates with the centroid, in the `geometry` of the spatial data frame. The package `sf` is necessary for the command `st_centroid`.
   
Additionally, the `annotate` function can be used to add a single character string at a specific location, as demonstrated here to add the "Gulf of Mexico"Pacific Ocean" and Ryukyu Archipelago

```{r,  eval=T}
library("sf")
world_points<- st_centroid(world)
world_points <- cbind(world, st_coordinates(st_centroid(world$geometry)))

ggplot(data = world) +
geom_sf() +
geom_text(data= world_points,aes(x=X, y=Y, label=name),
    color = "black", fontface = "bold", check_overlap = FALSE) +
    annotate(geom = "text", x = 124, y = 21, label = "Pacific Ocean", fontface = "italic", color = "#0b3c8a", size = 5) +
   annotate(geom = "text", x = 124.2, y = 24, label = "Ryukyu archipelago", fontface = "italic", color = "#d41919", size = 3) + 
   coord_sf(xlim = c(118, 128), ylim = c(17, 27), expand = FALSE)
```


- Finalization

Now to make the final touches, the theme of the map can be edited to make it more appealing. We suggested the use of `theme_bw` for a standard theme, but there are many other themes that can be selected from (see for instance `?ggtheme` in `ggplot2`, or the package `ggthemes` which provide several useful themes). Moreover, specific theme elements can be tweaked to get to the final outcome:

- Position of the legend: Although not used in this example, the argument `legend.position` allows to automatically place the legend at a specific location (e.g. "topright", "bottomleft", etc.);

- Grid lines (graticules) on the map: by using `panel.grid.major` and `panel.grid.minor`, grid lines can be adjusted. Here we set them to a gray color and dashed line type to clearly distinguish them from country borders lines;

- Map background: the argument `panel.background` can be used to color the background, which is the ocean essentially, with a light blue;
Many more elements of a theme can be adjusted, which would be too long to cover here. We refer the reader to the documentation for the function `theme`.

```{r,  eval=T}
ggplot(data = world) +
   geom_sf(fill= 'antiquewhite') + 
   geom_text(data= world_points,aes(x=X, y=Y, label=name), color = 'darkblue', fontface = 'bold', check_overlap = FALSE) + 
   annotate(geom = 'text', x = -90, y = 26, label = 'Gulf of Mexico', fontface = 'italic', color = 'grey22', size = 6) +
   annotation_scale(location = 'bl', width_hint = 0.5) + 
   annotation_north_arrow(location = 'bl', which_north = 'true', pad_x = unit(0.75, 'in'), pad_y = unit(0.5, 'in'), style = north_arrow_fancy_orienteering) + 
   coord_sf(xlim = c(-102.15, -74.12), ylim = c(7.65, 33.97), expand = FALSE) + 
   xlab('Longitude') + ylab('Latitude') + 
   ggtitle('Map of the Gulf of Mexico and the Caribbean Sea') + 
   theme(panel.grid.major = element_line(color = gray(.5), linetype = 'dashed', size = 0.5), panel.background = element_rect(fill = 'aliceblue'))
```

- Exporting map

The final map is now ready, it is very easy to save it using ggsave. This function allows a graphic (typically the last plot displayed) to be saved in a variety of formats, including the most common PNG (raster bitmap) and PDF (vector graphics), with control over the size and resolution of the outcome. For instance here, we save a PDF version of the map, which keeps the best quality, and a PNG version of it for web purposes:

```{r,  eval=T}
ggsave("Output/Datamap.pdf")
ggsave("Output/map_web.png", width = 6, height = 6, dpi = "screen")
```


### Species distribution 

```{r,  eval=T}
library(rgbif)
library(ggplot2)
library(mapr)

gbif.res <- occ_search(scientificName = "Urocissa caerulea", limit = 1200)

map_ggplot(gbif.res) +
  coord_sf(xlim = c(120, 123), ylim = c(21, 26), expand = FALSE)
```

Have a look [here](https://damariszurell.github.io/HU-GCIB/1_Data.html#2_environmental_data), to combine climate and species data. I have been trying to do it for few days but packages need API keys. This tutorial (not ggplot) should work - but not tested.

### Bathymetric map with  `marmap`

`marmap` can query and plot NOAA's bathymetry database

```{r,  eval=T}
library(marmap)
# download data
TW.bathy <- getNOAA.bathy(lon1=118,lon2=124, lat1=21,lat2=26,resolution=1) # don't put too wide / resolution: 1 
# define palette
blues <- colorRampPalette(c("darkblue", "cyan"))
greys <- colorRampPalette(c(grey(0.4),grey(0.99)))
# make the plot
plot.bathy(TW.bathy,
     image=T,
     deepest.isobath=c(-6000,-120,-30,0),
     shallowest.isobath=c(-1000,-60,0,0),
     step=c(2000,60,30,0),
     lwd=c(0.3,1,1,2),
     lty=c(1,1,1,1),
     col=c("grey","black","black","black"), 
     drawlabels=c(T,T,T,F),
     bpal = list(c(0,max(TW.bathy),greys(100)),c(min(TW.bathy),0,blues(100))),
     land=T, xaxs="i"
     )
```

Profiles can be extract using `get.transect`:

```{r,  eval=T}
tw.profile <-get.transect(TW.bathy,x1=119.5,y1=23.75, x2=122,y2=23.75, dis=TRUE)
plotProfile(tw.profile) 
#### Not Run: extract a profile Manually
#### manual.profile<-get.transect (TW.bathy, loc=T,dist=T) 
#### plotProfile(manual.profile)
```

### Interactive maps

```{r,  eval=T}
library(sf)
library(RColorBrewer)
library(dplyr)
library(devtools)
library(ggsflabel)
# devtools::install_github("yutannihilation/ggsflabel") #None

df.pop <- read.csv("Data/tw_population.csv", h = T)
df.pop <- data.frame(df.pop)

ggplot(df.pop, aes(x = reorder(County, Population), y = Population/10000, fill = Category)) +
  geom_bar(stat="identity") +
  coord_flip() +
  labs(title = "Taiwan Population", x = "County", y = "Population(萬)")

taiwan.map <- st_read ("Data/GADM/gadm36_TWN_2.shp")

ggplot(data = taiwan.map) +
  geom_sf() +
  labs(title = "Taiwan Map ")

ggplot(data = taiwan.map) +
  geom_sf(aes(fill = NAME_2), show.legend= F) +
  geom_sf_text(aes(label = NAME_2), size = 3) +
  labs(title = "Taiwan Map ")

ggplot(data = taiwan.map) +
  geom_sf(aes(fill = NAME_2)) +
  scale_fill_manual(name = "County",values = colorRampPalette(brewer.pal(8, "Accent"))(22)) +
  labs(title = "Taiwan Map ")
```

Combine the two datasets:


```{r,  eval=T}
my.taiwan.map <- taiwan.map[c("NAME_2", "geometry")]
my.taiwan.map$NAME_2 <- as.character(my.taiwan.map$NAME_2)
my.taiwan.map.data <- left_join(my.taiwan.map, df.pop,by= c("NAME_2" = "County"))

ggplot(data = my.taiwan.map.data) +
  geom_sf(aes(fill = Population))

ggplot(data = my.taiwan.map.data) +
  geom_sf(aes(fill = Population/10000)) +
  scale_fill_distiller(palette = "Spectral", name = "Population(x10,000)") +
  geom_sf_label_repel(aes(label = NAME_2, alpha = 1)) +
  labs(title="Taiwan popualtion map", x ="Longitude", y = "Latitude")
```

Make it interactive:

```{r,  eval=T}
# not run
# library(mapview)
# library(fields)
# mapview(my.taiwan.map.data["Population"])
# mapview(my.taiwan.map.data["Population"], col.regions = tim.colors(100))
```